Protein-intrinsic properties and epigenetic effects regulate pioneer-factor binding and function

### 2023-03-06 manuscript/manuscript.qmd
update MS draft
update abstract draft

---
title: "Defining properties regulating pioneer-factor binding and function"
author:
  - Tyler J. Gibson, Melissa M. Harrison
  # - Melissa M. Harrison
affiliations:
  - name: Department of Biomolecular Chemistry, University of Wisconsin School of Medicine and Public Health
    city: Madison
    state: WI
format:
  docx:
    reference-doc: custom-reference-doc.docx
editor: visual
bibliography: references.bib
csl: nsmb.csl
---

@@

<!-- # Abstract -->

<!-- {{< include manuscript_sections/_01_Abstract.qmd >}} -->

<!-- {{< pagebreak >}} -->

>>>>

---
title: "Protein-intrinsic properties and epigenetic effects regulate pioneer-factor binding and function"
author:
  - Tyler J. Gibson, Melissa M. Harrison
  # - Melissa M. Harrison
affiliations:
  - name: Department of Biomolecular Chemistry, University of Wisconsin School of Medicine and Public Health
    city: Madison
    state: WI
format:
  docx:
    reference-doc: custom-reference-doc.docx
editor: visual
bibliography: references.bib
csl: nsmb.csl
---

Department of Biomolecular Chemistry, University of Wisconsin School of Medicine and Public Health, Madison, WI

@@

# Abstract
{{< include manuscript_sections/_01_Abstract.qmd >}}

### 2023-03-06 manuscript/manuscript_sections/_01_Abstract.qmd
update MS draft

During development, transcription factors establish cell-type specific gene expression programs.
While many transcription factors cannot bind to target DNA in the context of nucleosomes, pioneer transcription factors access nucleosomal targets and promote chromatin opening.
Despite the ability to bind to nucleosomal target motifs, many pioneer factors display cell-type specific binding and activity.
It is unclear how pioneer factor occupancy is regulated in different cell types during development and how chromatin occupancy relates to chromatin opening activity.
To determine how pioneer-factor binding and activity are regulated, we systematically studied three Drosophila transcription factors with distinct DNA-binding domains and biological functions: Zelda, Grainy head, and Twist.
We demonstrated that the level of chromatin occupancy is a key determinant of pioneer-factor activity.
Zld, Grh and Twi occupancy are regulated by a numerous factors including motif content at binding sites, local chromatin context, and protein concentration.
We found that regions outside the DNA-binding domain are required for pioneer-factor binding and chromatin opening.
Together, our results demonstrate that both protein-intrinsic and cell-type specific factors regulate pioneer-factor occupancy and activity.
These findings have important implications for how pioneer factors drive development and how misexpression can lead to disease.

>>>>

Transcription factors establish cell-type specific gene expression programs to drive developmental transitions.
While chromatin presents a barrier to the binding of many transcription factors, pioneer factors access nucleosomal targets and promote chromatin opening.
Despite the ability to bind to target motifs in closed chromatin, many pioneer factors display cell-type specific binding and activity.
The mechanisms that govern pioneer-factor occupancy in different cell types and the relationship between chromatin occupancy and chromatin opening remain unclear.
To determine how pioneer-factor binding and activity are regulated, we systematically studied three Drosophila transcription factors with distinct DNA-binding domains and biological functions: Zelda, Grainy head, and Twist.
We demonstrated that the level of chromatin occupancy is a key determinant of pioneer-factor activity.
Zelda, Grainy head and Twist occupancy is regulated by multiple factors, including motif content at binding sites, local chromatin context, and protein concentration.
We found that regions outside the DNA-binding domain are required for binding and chromatin opening.
Together, our results demonstrate that pioneering activity is not a binary feature intrinsic to a protein but occurs on a spectrum and can be regulated by a variety of protein-intrinsic and cell-type-specific features.
These findings have important implications for how pioneer factors drive development and how misexpression can lead to disease.

### 2023-03-06 manuscript/manuscript_sections/_03_Results.qmd
update MS text
update MS draft

---
#bibliography: ../references.bib
---

@@

To determine how exogenous Zld and Grh interact with preexisting chromatin, we expressed each factor at approximately physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
Additional ChIP experiments were performed in WT cells using anti-Zld, or anti-Grh to control for antibody specificity.
By excluding peaks that were called in these datasets, we ensured that subsequent analysis was performed on peaks that were detected only upon ectopic expression of Zld or Grh.

@@

@@

@@

@@

@@

@@

Binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) [@Zaret2020].
However, little is known about how regions outside the DBD may contribute to pioneering function.
To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq on stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig. 10a,b).
Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was unable to bind closed chromatin at class II or III regions (Fig. 6a,c).
We previously showed that in neural stem cells, Zld binds to many regions that lack a canonical Zld motif through a mechanism that is not understood [@Larson2021a].
Because many Zld-binding sites in S2 cells also lack a canonical CAGGTA motif (Fig. 3), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding.
To rule out this possibility, we analyzed DBD ChIP-seq signal only at regions containing a canonical motif.
This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required (Extended Data Fig. 10a).
Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).
For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
Together, these experiments show that regions outside the DBD are essential for binding and opening closed chromatin.

>>>>

---
# bibliography: ../references.bib
---

@@

To determine how exogenous Zld and Grh interact with preexisting chromatin, we expressed each factor at approximately physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
Additional ChIP experiments were performed in WT cells using anti-Zld, or anti-Grh to control for antibody specificity.
By using these datasets as controls for peak calling, we ensured that subsequent analysis was performed on peaks that were detected only upon ectopic expression of Zld or Grh (Extended Data Fig. 1e-g).

@@

::: {#fig-1}
![](figures/fig1.pdf)

**Ectopic pioneer factors open chromatin and activate transcription. a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for WT cells lacking Zld or Grh expression.
**c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**.
**d,g,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions for Zld **(d)** or Grh **(g)**.
**e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h). i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
**j,i,** Proportion of class I, II or III regions that are proximal to a gene that is upregulated upon Zld **(j)** or Grh **(i)** expression.
Promoter-proximal and promoter-distal binding sites are shown separately.
The grey dotted line indicates the percentage of all binding sites that are proximal to a upregulated expressed gene.
:::

@@

::: {#fig-2}
![](figures/fig2.pdf)

**Twist binds closed chromatin extensively and drives accessibility at a limited number of sites. a,** Genome browser tracks showing examples of individual class I, II or III regions **b,** Pie charts showing the distribution of class I, II, and III binding sites.
**c,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions.
**d,** Heatmap showing ChIP-seq and ATAC-seq signal at class II and III regions.
:::

@@

::: {#fig-3}
![](figures/fig3.pdf)

**Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
**b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.
Zld, Grh, and Twi ChIP-seq data were generated in this study.
Remaining ChIP-seq datasets were previously published (see Supplementary table XX).
**c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
Logos plot of the canonical motif are shown alongside the heatmap.
Percentages are shown for class I, II or III binding sites, or all WT ATAC-seq peaks as a control for the background motif frequency.
**f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**.
**g,i,k,** Boxplots of the average motif log-odds score for class I, II or III regions for Zld **(g)**, Grh **(i)**, or Twi **(k)**.
Motif scores were determined by comparing each identified motif instance to the canonical motif.
:::

@@

::: {#fig-4}
![](figures/fig4.pdf)

**Many endogenous binding sites are resistant to ectopic pioneer-factor binding. a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
ChIP-seq data for repressive histone modifications H3K27me3 and H3K9me3 is also shown.
**d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
Zld, Grh, and Twi ChIP-seq data are z-score normalized.
CUT&RUN data for H3K27me3 is spike-in normalized using barcoded nucleosomes (see methods).
:::

@@

::: {#fig-5}
![](figures/fig5.pdf)

**Binding and opening of closed chromatin is concentration dependent. a,e,i,** Metaplots of ChIP-seq signal at class I-VI regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
Concetrations of CuSO~4~ used to induce protein expression are indicated above each plot.
**b,f,j,** Metaplots of ATAC-seq signal at class I-VI regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
**c,g,k,** Distribution of class I, II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
Y-axis indicates the total number of peaks called at each concentration.
Colors indicate the proportion of those peaks that are defined as class I, II, or III.
**d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.
:::

@@

Binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) [@Zaret2020].
However, little is known about how regions outside the DBD may contribute to pioneering function.
We generated stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig. 10a,b).
Immunostaining confirmed that the each DBD alone was properly localized to the nucleus (Extended Data Fig. 10c-d) To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq.
Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was unable to bind closed chromatin at class II or III regions (Fig. 6a,c).
We previously showed that in neural stem cells, Zld binds to many regions that lack a canonical Zld motif through a mechanism that is not understood [@Larson2021a].
Because many Zld-binding sites in S2 cells also lack a canonical CAGGTA motif (Fig. 3), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding.
To rule out this possibility, we analyzed DBD ChIP-seq signal only at regions containing a canonical motif.
This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required (Extended Data Fig. 10a).
Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).
For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
Together, these experiments show that regions outside the DBD are essential for binding and opening closed chromatin.

::: {#fig-6}
![](figures/fig6.pdf)

**Regions outside the DNA-binding domain are required for pioneer-factor function. a-b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
Shown for Zld **(a)** and Grh **(b)**.
**c,g,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(g)**.
**d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
:::

### 2023-03-06 manuscript/manuscript_sections/_04_Discussion.qmd
update MS draft

---
format: html
editor: visual
#bibliography: ../references.bib
---

In this study, we systematically explore chromatin binding and opening by the transcription factors Zld, Grh and Twi.
Unlike many mammalian TFs, which have mostly been studied in cell culture, Zld, Grh and Twi have been extensively studied during normal Drosophila development.
By comparing the results of TF induction in S2 cells to Zld, Grh and Twi activity during normal development, we are able to provide mechanistic insights that are relevant to the endogenous function of pioneer factors.
Zld, Grh, and Twi displayed a range of abilities to bind and open closed chromatin, suggesting that pioneering activity occurs on a spectrum and is context dependent.
Binding is separable from chromatin opening through the identification of class II sites that are bound by PFs but remain closed, or by the expression of the Grh DBD, which retains some binding but cannot promote chromatin opening.
The separability of binding and opening suggests that additional downstream steps such as recruitment of chromatin remodelers are required for chromatin opening [@Judd2021; @King2017].
Local restructuring of nucleosome-DNA interactions [@Michael2020; @Dodonova2020], if it occurs in the case of Zld, Grh, or Twi, may be insufficient for establishing NFRs and may instead stabilize PF binding to allow recruitment of cofactors.
We propose that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features (Fig 7) such as local chromatin features, protein concentration, DNA motif content, and protein domains outside the DBD.

Despite the ability of PFs to bind to closed chromatin, most PFs occupy only a subset of target motifs in the genome and PF binding and activity is often cell-type specific [@Larson2021].
In many cases, the presence of active or repressive chromatin is an important determinant of transcription factor binding, but the extent to which this informs PF binding is unclear.
For the reprogramming factors Oct4, Sox2 and Klf4, and c-Myc (OSKM), high levels of the repressive histone modification H3K9me3 mark regions that are resistant to OSKM binding.
Depletion of H3K9 methyltransferases increases OSKM occupancy at these regions [@Soufi2012].
In the case of Pax7, H3K27me3 is correlated with cell-type specific Pax7 binding [@Mayran2019].
Zld, Grh and Twi binding showed similar correlations with local chromatin structure.
Even when expressed at low levels, all three proteins bind to active, accessible regions with degenerate motifs or no detectable motifs at all.
At approximately physiological levels, these factors bind to canonical motifs in naïve chromatin, and are largely excluded from repressed chromatin (bearing H3K9me3 or H3K27me3).
When we depleted H3K27me3 with the polycomb inhibitor tazemetostat, we observed no effect on binding within these repressed regions.
In contrast to taz treatment, expression of Zld and Grh, but not Twi, at higher levels allowed at least some binding to these repressed regions.
These results suggest that despite the correlation with H3K27me3, protein concentration and affinity for target motifs may be more important than the presence of polycomb silencing for tissue-specific binding and emphasize the importance of functionally testing the importance of different chromatin marks to PF binding.
Cooperative binding with cell-type specific cofactors can also stabilize PF binding to tissue-specific binding sites [@Donaghey2018].
This may also be important for Zld, Grh, and Twi, and may explain the presence of class VI sites that are resistant to binding and lack any repressive histone modifications.
Further exploration of how chromatin structure and cofactor expression influence PF binding and activity will be essential to developing a more complete understanding of how these factors function in development and disease.

The strong concentration dependence of chromatin opening by Zld, Grh and Twi provides important context for the *in vivo* biological functions of these proteins.
Zld, along with many other TFs, forms high concentration foci within the nucleus [@Mir2018].
These foci may facilitate chromatin opening by pioneer factors by creating a high protein concentration at specific genomic regions or locations within the nucleus.
Grh has pioneering activity in late drosophila embryos and larval tissues, but not in early embryos [@Nevil2020; @Jacobs2018].
This could, in part, be accounted for by differences in Grh expression between tissues.
Additionally, GRHL2, one the mammalian Grh orthologs, can act as either a tumor suppressor or an oncogene, depending on the context [@Reese2019].
Considering Grh expression levels in different tissues, cell types, and cancer types will be important for determining these contradictory effects.
This is especially relevant considering that GRHL2 is frequently subject to copy number amplification in various cancers, leading to overexpression [@Reese2019].
Pioneering activity by other mammalian factors depends on protein concentration [@Hansen2022; @Blassberg2022], suggesting that this is likely a general feature of PFs.

@@

>>>>

---
format: html
editor: visual
# bibliography: ../references.bib
---

In this study, we systematically explore chromatin binding and opening by the transcription factors Zld, Grh and Twi.
Unlike many mammalian TFs, which have mostly been studied in cell culture, Zld, Grh and Twi have been extensively studied during normal Drosophila development.
By comparing the results of TF induction in S2 cells to Zld, Grh and Twi activity during normal development, we are able to provide mechanistic insights into how PFs are regulated in different cell types, and how PF binding relates to function.
Zld, Grh, and Twi displayed a range of abilities to bind and open closed chromatin, suggesting that pioneering activity occurs on a spectrum and is context dependent.
Binding is separable from chromatin opening through the identification of class II sites that are bound by PFs but remain closed, or by the expression of the Grh DBD, which retains some binding but cannot promote chromatin opening.
The separability of binding and opening suggests that additional downstream steps such as recruitment of chromatin remodelers are required for chromatin opening [@Judd2021; @King2017].
Local restructuring of nucleosome-DNA interactions [@Michael2020; @Dodonova2020], if it occurs in the case of Zld, Grh, or Twi, may be insufficient for establishing accessible chromatin and may instead stabilize PF binding to allow recruitment of cofactors.
We find that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features such as local chromatin features, protein concentration, DNA motif content, and protein domains outside the DBD.

Despite the ability of PFs to bind to closed chromatin, most PFs occupy only a subset of target motifs in the genome and PF binding and activity is often cell-type specific [@Larson2021].
In many cases, the presence of active or repressive chromatin is an important determinant of transcription factor binding, but the extent to which this informs PF binding is unclear.
For the reprogramming factors Oct4, Sox2 and Klf4, and c-Myc (OSKM), high levels of the repressive histone modification H3K9me3 mark regions that are resistant to OSKM binding.
Depletion of H3K9 methyltransferases increases OSKM occupancy at these regions [@Soufi2012].
In the case of Pax7, H3K27me3 is correlated with cell-type specific Pax7 binding [@Mayran2019].
Zld, Grh and Twi binding showed similar correlations with local chromatin structure.
Even when expressed at low levels, all three proteins bind to active, accessible regions with degenerate motifs or no detectable motifs at all.
At approximately physiological levels, these factors bind to canonical motifs in naïve chromatin, and are largely excluded from repressed chromatin (bearing H3K9me3 or H3K27me3).
When we depleted H3K27me3 with the polycomb inhibitor tazemetostat, we observed no effect on binding within these repressed regions.
In contrast to taz treatment, expression of Zld and Grh, but not Twi, at higher levels allowed at least some binding to these repressed regions.
These results suggest that despite the correlation with H3K27me3, protein concentration and affinity for target motifs may be more important than the presence of polycomb silencing for tissue-specific.
Cooperative binding with cell-type specific cofactors can also stabilize PF binding to tissue-specific binding sites [@Donaghey2018].
This may also be important for Zld, Grh, and Twi, and may explain the presence of class VI sites that are resistant to binding and lack any repressive histone modifications.
Further exploration of how chromatin structure and cofactor expression influence PF binding and activity will be essential to developing a more complete understanding of how these factors function in development and disease.

@@

The strong concentration dependence of chromatin opening by Zld, Grh and Twi provides important context for the *in vivo* biological functions of these proteins.
Zld, along with many other TFs, forms high concentration foci within the nucleus [@Mir2018].
These foci may facilitate chromatin opening by pioneer factors by creating a high protein concentration at specific genomic regions or locations within the nucleus.
Grh has pioneering activity in late drosophila embryos and larval tissues, but not in early embryos [@Nevil2020; @Jacobs2018].
This could, in part, be accounted for by differences in Grh expression between tissues.
Pioneering activity by other mammalian factors depends on protein concentration [@Hansen2022; @Blassberg2022], suggesting that this is likely a general feature of PFs.

Due to the essential role of PFs in regulating cell fate, their misexpression frequently results in disease [@Larson2021; @Dobersch2019].
GRHL2, one the mammalian Grh orthologs, is frequently subject to copy number amplification in various cancers, leading to overexpression, and can act as either a tumor suppressor or an oncogene, depending on the context [@Reese2019].
Misexpression of the PF DUX4 results in the muscular atrophy disease facioscapulohumeral muscular dystrophy (FSHD) [@Campbell2018].
Our data provide evidence for how misexpression and overexpression of PFs can lead to ectopic activation of gene-regulatory programs.
A better understanding of how PF binding and activity is regulated will be essential for understanding and treating diseases caused by these proteins.
